On exponential sums arising from the classical doubling method

### 2025-02-28 paper.tex
Even orthogonal groups
Defining the extension only for cases needed

Let $\hermitianSpace$ be a vector space of dimension $n$ over $\quadraticExtension$, equipped with a non-degenerate sesquilinear form $\innerproduct{\cdot}{\cdot} \colon \hermitianSpace \times \hermitianSpace \to \quadraticExtension$ which is $\epsilon$-symmetric for $\epsilon \in \left\{\pm 1\right\}$. By this we mean that:
\begin{enumerate}
	\item For every $x_1,x_2,y \in \hermitianSpace$, $$\innerproduct{x_1 + x_2}{y} = \innerproduct{x_1}{y} + \innerproduct{x_2}{y}.$$
	\item For every $x,y \in \hermitianSpace$ and $t \in \quadraticExtension$, $$\innerproduct{tx}{y} = t\innerproduct{x}{y}.$$
	\item ($\epsilon$-symmetric) For every $x,y \in \hermitianSpace$, $$\innerproduct{x}{y} = \epsilon \involution{\innerproduct{y}{x}}.$$
	\item (non-degenerate) For every $0 \ne x \in \hermitianSpace$ there exists $y \in \hermitianSpace$ such that $$\innerproduct{x}{y} \ne 0.$$
\end{enumerate}
Let $\IsometryGroup \left(\hermitianSpace\right)$ be the isometry group of $\hermitianSpace$, consisting of all the elements of $\GL\left(\hermitianSpace\right)$ that satisfy $\innerproduct{gx}{gy} = \innerproduct{x}{y}$ for every $x,y \in \hermitianSpace$. We denote by $G$ the identity component of $\IsometryGroup\left(\hermitianSpace\right)$. Let $\GroupExtension{\IsometryGroup\left(\hermitianSpace\right)}$ be the similitude group version of $\IsometryGroup\left(\hermitianSpace\right)$, consisting of the elements $g \in \GL\left(\hermitianSpace\right)$ with the property that there exists a constant $\similitudeCharacter\left(g\right) \in \multiplicativegroup{\quadraticExtension}$ such that for every $\innerproduct{gx}{gy} = \similitudeCharacter\left(g\right) \innerproduct{x}{y}$ for every $x,y \in \hermitianSpace$. Let $\GroupExtension{G}$ be the identity component of $\GroupExtension{\IsometryGroup\left(\hermitianSpace\right)}$. Let $\lieAlgebra$ be the Lie algebra of $G$, consisting of all elements $A \in \EndomorphismRing\left(\hermitianSpace\right)$ satisfying $\innerproduct{Ax}{y} + \involution{\innerproduct{x}{AY}} = 0$ for every $x, y \in \hermitianSpace$.

@@

Let $\pi$ be an irreducible representation of $H$ where $H \in \{G, \GroupExtension{G}\}$. Denote $$\dblGaussSum{\pi}{\chi} = \frac{1}{\sqrt{\sizeof{\lieAlgebra}}} \sum_{g \in G} \pi\left(g\right) \Phi_{\chi}\left(g\right).$$
Since $\Phi_{\chi}$ is a class function of $\GL\left(\hermitianSpace\right)$, it is also a class function of $G$. Since $G$ is a normal subgroup of $H$, we have that $\dblGaussSum{\pi}{\chi} \in \Hom_{H}\left(\pi, \pi\right)$. By Schur's lemma, there exists a constant $\dblGaussSumScalar{\pi}{\fieldCharacter} \in \cComplex$, such that $$\dblGaussSum{\pi}{\chi} = \dblGaussSumScalar{\pi}{\chi} \cdot \idmap_\pi.$$
We call $\dblGaussSumScalar{\pi}{\chi}$ the \emph{doubling method Gauss sum}.

>>>>

Let $\hermitianSpace$ be a vector space of dimension $n$ over $\quadraticExtension$, equipped with a non-degenerate sesquilinear form $\innerproduct{\cdot}{\cdot} \colon \hermitianSpace \times \hermitianSpace \to \quadraticExtension$ which is $\epsilon$-symmetric for $\epsilon \in \left\{\pm 1\right\}$. By this we mean that:
\begin{enumerate}
	\item For every $x_1,x_2,y \in \hermitianSpace$, $$\innerproduct{x_1 + x_2}{y} = \innerproduct{x_1}{y} + \innerproduct{x_2}{y}.$$
	\item For every $x,y \in \hermitianSpace$ and $t \in \quadraticExtension$, $$\innerproduct{tx}{y} = t\innerproduct{x}{y}.$$
	\item ($\epsilon$-symmetric) For every $x,y \in \hermitianSpace$, $$\innerproduct{x}{y} = \epsilon \involution{\innerproduct{y}{x}}.$$
	\item (non-degenerate) For every $0 \ne x \in \hermitianSpace$ there exists $y \in \hermitianSpace$ such that $$\innerproduct{x}{y} \ne 0.$$
\end{enumerate}
Let $\IsometryGroup \left(\hermitianSpace\right)$ be the isometry group of $\hermitianSpace$, consisting of all the elements of $\GL\left(\hermitianSpace\right)$ that satisfy $\innerproduct{gx}{gy} = \innerproduct{x}{y}$ for every $x,y \in \hermitianSpace$. We denote by $G$ the identity component of $\IsometryGroup\left(\hermitianSpace\right)$. Let $\GroupExtension{\IsometryGroup\left(\hermitianSpace\right)}$ be the similitude group version of $\IsometryGroup\left(\hermitianSpace\right)$, consisting of the elements $g \in \GL\left(\hermitianSpace\right)$ with the property that there exists a constant $\similitudeCharacter\left(g\right) \in \multiplicativegroup{\quadraticExtension}$ such that for every $\innerproduct{gx}{gy} = \similitudeCharacter\left(g\right) \innerproduct{x}{y}$ for every $x,y \in \hermitianSpace$. Let $\GroupExtension{G}$ be the identity component of $\GroupExtension{\IsometryGroup\left(\hermitianSpace\right)}$. If $\quadraticExtension = \finiteField$ and $\dim_{\finiteField} \hermitianSpace = 2n$ is even, then $\GroupExtension{G}$ consists of all element in $G$ satisfying $\similitudeCharacter\left(g\right) = \left(\det g\right)^n$. Let $\lieAlgebra$ be the Lie algebra of $G$, consisting of all elements $A \in \EndomorphismRing\left(\hermitianSpace\right)$ satisfying $\innerproduct{Ax}{y} + \involution{\innerproduct{x}{AY}} = 0$ for every $x, y \in \hermitianSpace$.

We will assume that $\dim_{\finiteField} \hermitianSpace$ is even. This is automatic for all case, expect for when $\quadraticExtension = \finiteField$ and $\epsilon = 1$, in which case we will only consider even special orthogonal groups.

@@

Let $\pi$ be an irreducible representation of $H$ where $H \in \{G, \GroupExtension{G}\}$ if $\quadraticExtension = \finiteField$ and $H = G$ if $\quadraticExtension \ne \finiteField$. Denote $$\dblGaussSum{\pi}{\chi} = \frac{1}{\sqrt{\sizeof{\lieAlgebra}}} \sum_{g \in G} \pi\left(g\right) \Phi_{\chi}\left(g\right).$$
Since $\Phi_{\chi}$ is a class function of $\GL\left(\hermitianSpace\right)$, it is also a class function of $G$. Since $G$ is a normal subgroup of $H$, we have that $\dblGaussSum{\pi}{\chi} \in \Hom_{H}\left(\pi, \pi\right)$. By Schur's lemma, there exists a constant $\dblGaussSumScalar{\pi}{\fieldCharacter} \in \cComplex$, such that $$\dblGaussSum{\pi}{\chi} = \dblGaussSumScalar{\pi}{\chi} \cdot \idmap_\pi.$$
We call $\dblGaussSumScalar{\pi}{\chi}$ the \emph{doubling method Gauss sum}.
